---
title: "Activity Monitoring"
author: "Gerald Yong"
date: "04/19/2015"
output: 
  html_document:
    pandoc_args: [
      "+RTS", "-K64m", "-RTS"
    ]
---

### Loading and Pre-processing

1. Download the data from the web.

```{r}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
temp <- tempfile()
download.file(fileURL, destfile=temp, method="curl")
con <- unz(temp, "activity.csv")
```

2. Load the data from CSV into a data frame and view the top 15 rows.

```{r}
activitydata <- read.csv(con, header=TRUE, na.strings=c("NA")) 
head(activitydata, 15)
unlink(temp)
```

3. Perform some formatting of the columns into the right data types.

```{r}
library(lubridate)
activitydata$steps <- as.numeric(activitydata$steps)
activitydata$interval <- as.numeric(activitydata$interval)
activitydata$date <- ymd(activitydata$date)
str(activitydata)
```

4. Load other required libaries.

```{r message=FALSE, warning=FALSE}
library(dplyr)
```

### What is the mean total number of steps taken per day?

1. Calculate the total number of steps taken per day.

```{r}
ad <- activitydata %>%
        group_by(date) %>%
        filter(!is.na(steps)) %>%
        summarise(sumsteps = sum(steps)) %>%
        print()
```

2. Generate the histogram to show the frequency distribution of the number of steps.

```{r}
hist(ad$sumsteps)
```

3. Determine the mean and median of the total number of steps.

```{r}
meansteps <- mean(ad$sumsteps)
medsteps <- median(ad$sumsteps)
meansteps
medsteps
```

The mean of the daily total number of steps is **`r round(meansteps,2)`** and the median is **`r round(medsteps,2)`**.

### What is the average daily pattern?

1. Determine the average number of steps for each interval across all days.

```{r}
ad <- activitydata %>%
        group_by(interval) %>%
        filter(!is.na(steps)) %>%
        summarise(sumsteps = sum(steps), 
                  meansteps = mean(steps), 
                  mediansteps = median(steps)) %>%
        print()
str(ad)
```

2. Create a time-series plot.
```{r}
plot(y=ad$meansteps, x=ad$interval, type="l")
```

3. Determine the maximum number of steps and the interval it belongs in.

```{r}
maxsteps <- max(ad$meansteps)
whichint <- ad[ad$meansteps==maxsteps,]$interval
maxsteps
whichint
```

The maximum number of steps is **`r maxsteps`** and it occurred in interval **`r whichint`**.

### Imputing missing values

1. Count the number of NAs.

```{r}
numNA <- sum(ifelse(is.na(activitydata$steps),1,0))
numNA
```

There are **`r numNA`** NA records.

2. To fill up the NAs, assign it to the median no. of steps for the interval.

```{r}
ad2 <- activitydata
ad2$steps <- apply(ad2, 1, function(x) ifelse(is.na(x[1]),ad[ad$interval==as.numeric(x[3]),]$mediansteps,x[1]))
ad2$steps <- as.numeric(ad2$steps)
str(ad2)
```

3. List out a portion of the new data set.

```{r}
head(ad2)
```

4. Generate the histogram.

```{r}
hist(ad2$steps)
```

5. Determine the mean and median.

```{r}
ad3 <- ad2 %>%
        group_by(date) %>%
        summarise(sumsteps = sum(steps),
                  meansteps = mean(steps), 
                  mediansteps = median(steps)) %>%
        print()

meansteps2 <- mean(ad3$sumsteps)
medsteps2 <- median(ad3$sumsteps)
```

The mean and median number of steps is **`r meansteps2`** and **`r medsteps2`** respectively. We can do up a table to compare the original and the imputed means and medians.

```{r}
tblcompare <- matrix(c(meansteps, medsteps, meansteps2, medsteps2, 
                       (meansteps2-meansteps)/meansteps*100, 
                       (medsteps2-medsteps)/medsteps*100), 
                       nrow=2, ncol=3)
colnames(tblcompare) <- c("orig","imputed", "%-change")
rownames(tblcompare) <- c("mean", "median")
tblcompare
```

It can be seen that imputing values resulted in a lower median and mean.

### Are there differences in activity patterns between weekdays and weekends

1. Create the new weekdaytype factor variable.

```{r}
ad4 <- ad2 %>%
          mutate(weekdaytype = 
                   ifelse(weekdays(date,abbreviate=TRUE)=="Sat" |
                          weekdays(date,abbreviate=TRUE)=="Sun",
                          "weekend","weekday")) %>%
          group_by(date, weekdaytype) %>%
          summarise(sumsteps = sum(steps)) %>%
          print()
```

The above illustrates that the weekdaytype factor is working.

2. Aggregate by interval and weekday.

```{r}
ad5 <- ad2 %>%
          mutate(weekdaytype = 
                   ifelse(weekdays(date,abbreviate=TRUE)=="Sat" |
                          weekdays(date,abbreviate=TRUE)=="Sun",
                          "weekend","weekday")) %>%
          group_by(interval, weekdaytype) %>%
          summarise(sumsteps = sum(steps),
                    meansteps = mean(steps), 
                    mediansteps = median(steps)) %>%  
          print()
```

3. Plot average number of steps taken for each 5-min interval for all weekdays.

```{r message=FALSE}
library(ggplot2)
graph <- ggplot(ad5, aes(x=interval, y=meansteps)) +
  geom_line() +
#  geom_text(aes(size=1)) +
  facet_grid(weekdaytype ~ .) +
  labs(x="Interval", y="Avg No. of Steps") +
  ggtitle("Average No Of Steps By Interval and Weekday Type")
print(graph)
```